Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-05 _tex/index.tex
Built site for gh-pages

In its most general form, Thurstone's theory deals with pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. The theory proposes that two key
factors determine the dichotomous outcome of these comparisons: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} represents the psychological impact each
stimulus has on judges or, more simply, their underlying perception of
the stimulus' trait level. According to the theory, the discriminal
process for each stimulus follows a Normal distribution, where its mode
(mean), referred to as the \emph{modal discriminal process}, indicates
the stimulus' position on the trait continuum, and its dispersion, known
as the \emph{discriminal dispersion}, reflects the variability in the
perceived trait level of the stimulus.

@@

However, because the discriminal process of a single stimulus is not
directly observable, the theory introduces the \emph{law of comparative
judgment}. This law posits that in pairwise comparisons, a judge
perceives the stimulus positioned further along the trait continuum as
having a higher level of that trait. This principle highlights that the
outcome of a pairwise comparison likely depends on the relative distance
between stimuli rather than their absolute positions on the trait
continuum.

@@

Figure~\ref{fig-discriminal_difference} shows the distribution of the
discriminal difference for the texts depicted in
Figure~\ref{fig-discriminal_process}, assuming a correlation of
\(\rho = 0.6\). The figure reveals that, under these conditions, the
judge perceives text B as having significantly higher quality than text
A, as indicated by the shaded gray area under the curve \(P(B > A)\). As
a result, the dichotomous outcome of this comparison would mostly favor
text B over text A.

@@

Figure~\ref{fig-correlation} illustrates how varying correlations
influence the distribution of the discriminal difference for the texts
depicted in Figure~\ref{fig-discriminal_process}. Since the texts differ
in quality, higher correlations increase the likelihood that the
discriminal difference distinctly favors text B over text A. This is
evident from the larger proportion of the area under the curve,
\(P(B > A)\), that lies above zero (shaded gray area in
Figure~\ref{fig-correlation}) compared to curves with lower
correlations. Moreover, although the figure does not illustrate this
scenario, it is reasonable to infer that if the texts had similar or
identical quality levels, higher correlations would likely reduce the
chance of the discriminal difference distinctly favoring one text over
the other. This probability reduction occurs because the distribution of
the discriminal difference would become more narrowly centered around
zero.

@@

On the one hand, the assumption of \emph{equal dispersion between
stimuli} suggests that variability in perceived trait levels is
consistent across all stimuli. However, Thurstone observed that this
assumption may not hold when researchers assess complex traits or
heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
and stimuli can introduce greater judgment discrepancies due to their
unique characteristics. In fact, research on text quality may already
provide evidence of the violation of this assumption, particularly
through the use of misfit statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
Misfit statistics quantify the judgment discrepancies associated with a
given stimulus
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
In this context, texts identified as ``misfits'' display greater
judgment discrepancies than other texts. This implies that the
discriminal difference associated with these texts shows a broader
dispersion, suggesting that their discriminal processes also exhibit
greater variation than other texts. This reasoning also applies to
``misfit'' judges, whose evaluations reflect substantial deviations due
to the unique or extreme characteristics of the stimuli.

{what are the challenges when the equal dispersion assumption is
violated?}

@@

Fortunately, the same literature offers solutions for addressing these
issues. \citet{Andrich_1978} and \citet{Wainer_et_al_1978} recommend
integrating judges' biases into the BTL model. Moreover, the literature
advocates for the incorporation of relevant hierarchical structures into
the statistical model to account for these dependencies. Together, these
additions can result in a model resembling a Multilevel Structural
Equation Model (MSEM) \citep[chap.~26]{Hoyle_et_al_2023} combined with a
multidimensional or two-parameter logistic IRT model
\citep[chap.~15]{Hoyle_et_al_2023}, depending on the theoretical and
statistical treatment of judges' biases.

@@

Building on the previous section, it is evident that the BTL model
commonly functions as the trait's measurement model in CJ experiments
\citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
manifest variables contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
researchers use the BTL model to process the dichotomous outcomes
resulting from the pairwise comparisons (the manifest variables) to
estimate scores that reflect the underlying quality level of texts (the
latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

To mitigate these risks, principles from Structural Equation Modeling
(SEM) \citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}
recommend conducting these analyses and tests within a structural model.
A structural model specifies how different manifest or latent variables
influence the latent variable of interest \citep{Everitt_et_al_2010}.
This approach allows analyses that can account for both the BTL scores
and their uncertainties simultaneously, rather than treating them as
separate elements. Therefore, an integrated approach that combines CJ's
measurement and structural models can offer significant advantages.

>>>>

In its most general form, Thurstone's theory deals with pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. The theory proposes that two key
factors determine the dichotomous outcome of these comparisons: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} represents the psychological impact each
stimulus has on judges or, more simply, their underlying perception of
the stimulus' trait. According to the theory, the discriminal process
for each stimulus follows a Normal distribution, where its mode (mean),
referred to as the \emph{modal discriminal process}, indicates the
stimulus' position on the trait continuum, and its dispersion, known as
the \emph{discriminal dispersion}, reflects the variability of the
stimulus' perceived trait.

@@

However, because the discriminal process of a single stimulus is not
directly observable, the theory introduces the \emph{law of comparative
judgment}. This law posits that in pairwise comparisons, a judge
perceives the stimulus positioned further along the trait continuum as
having a higher level of that trait. This emphasizes that the outcome of
a pairwise comparison likely depends on the relative distance between
stimuli rather than their absolute positions on the trait continuum.

@@

Figure~\ref{fig-discriminal_difference} shows the distribution of the
discriminal difference for the texts depicted in
Figure~\ref{fig-discriminal_process}, assuming a correlation of
\(\rho = 0.6\). The figure reveals that, under these conditions, the
judge perceives text B as having significantly higher quality than text
A, as indicated by the shaded gray area under the curve \(P(B > A)\). As
a result, the dichotomous outcome of this comparison would likely favor
text B over text A.

@@

Figure~\ref{fig-correlation} illustrates how varying correlations
influence the distribution of the discriminal difference for the texts
depicted in Figure~\ref{fig-discriminal_process}. Since the texts differ
in quality, higher correlations increase the likelihood that the
discriminal difference distinctly favors text B over text A. This is
evident from the larger proportion of the area under the curve,
\(P(B > A)\), that lies above zero compared to curves with lower
correlations. Moreover, although the figure does not illustrate this
scenario, it is reasonable to infer that if the texts had similar or
identical quality levels, higher correlations would likely reduce the
chance of the discriminal difference distinctly favoring one text over
the other. This probability reduction occurs because the distribution of
the discriminal difference would become more narrowly centered around
zero.

@@

On the one hand, the assumption of \emph{equal dispersion between
stimuli} suggests that the variability of the perceived trait remains
consistent across all stimuli. However, Thurstone contended that this
assumption may not hold when researchers assess complex traits or
heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
and stimuli can introduce judgment discrepancies due to their unique
features. Indeed, evidence of such violation may already exist in the CJ
literature as misfit statistics. These statistics measure the judgment
discrepancies associated with a given stimulus
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For instance, labeling texts as ``misfits'' indicates that comparisons
involving these texts result in more judgment discrepancies than
comparisons involving other texts
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
This finding implies that the discriminal differences associated with
``misfits'' texts usually display a broader dispersion, suggesting that
their discriminal processes also exhibit more variation than other
texts. Notably, this reasoning also applies to ``misfit'' judges, whose
evaluations reflect substantial deviations from the shared consensus due
to the unique characteristics of the stimuli or the judges themselves.

However, assuming equal dispersions between stimuli despite its
violation can lead Case V (and the BTL model) to overlook critical
differences in the reliability of the trait across stimuli. Furthermore,
if researchers acknowledge that misfit statistics help identify these
critical differences, the common practice in the CJ literature of
excluding stimuli based on these statistics
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
discarding valuable information and introducing bias into the trait's
estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases are unpredictable, as they
depend on the particular stimuli researchers exclude from the analysis.
Taken together, these oversights undermine the reliability of the trait
and ultimately compromise its validity \citep[pp.~2]{Perron_et_al_2015}.

@@

Fortunately, the same literature offers solutions for addressing these
issues. \citet{Andrich_1978} and \citet{Wainer_et_al_1978} recommend
integrating judges' biases into the BTL model. Moreover, the literature
advocates for the incorporation of relevant hierarchical structures into
the statistical model to account for these dependencies. Together, these
additions can result in a model resembling a Multilevel Structural
Equation Model (MSEM) \citep[chap.~26]{Hoyle_et_al_2023} combined with a
multidimensional or two-parameter logistic Item Response Theory (IRT)
model \citep[chap.~15]{Hoyle_et_al_2023}, depending on the theoretical
and statistical treatment of judges' biases.

@@

Building on the previous section, it is evident that, besides its
inconveniences, the BTL model commonly functions as the trait's
measurement model for CJ data \citep{Andrich_1978, Bramley_2008}. A
measurement model specifies how manifest variables contribute to the
estimation of latent variables \citep{Everitt_et_al_2010}. For example,
when evaluating text quality, researchers use the BTL model to process
the dichotomous outcomes resulting from the pairwise comparisons (the
manifest variables) to estimate scores that reflect the underlying
quality level of texts (the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

To mitigate these risks, principles from Structural Equation Modeling
(SEM) \citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}
recommend conducting these analyses and tests within a structural model.
A structural model specifies how different manifest or latent variables
influence the latent variable of interest \citep{Everitt_et_al_2010}.
This approach allows analyses that can account for both the BTL scores
and their uncertainties simultaneously, rather than treating them as
separate elements.
